Genetic draft and valley crossing

### 2018-08-01 ms/valley_crossing.tex
update to Genetics template
minor typos

\documentclass[rmp]{revtex4}
\usepackage[english]{babel}
\usepackage{amssymb,amsfonts,amsmath}
\usepackage{xcolor}
\usepackage{breqn}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup{compatibility=false}

\usepackage{amsmath}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{enumerate}
%\usepackage{subfig}

\usepackage{fontspec}

% Code syntax highlighting
\usepackage{listings}
\lstloadlanguages{C++,Python}

@@

\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.42}

@@

\usepackage{setspace}
\usepackage[colorinlistoftodos]{todonotes}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Genetic draft and valley crossing}
\author{Taylor~Kessinger}
\author{Jeremy~Van~Cleve}
%\affiliation{Department of Biology, University of Kentucky}

\date{\today}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Living systems are characterized by complex adaptations which require multiple coordinated mutations in order to function. Empirical studies of fitness landscapes that result from the many possible mutations in a gene region reveal many fitness peaks and valleys that connect them. Thus, it is possible that some complex adaptations have arisen by evolutionary paths whose intermediate states are neutral or even deleterious.
When intermediates are deleterious, traversing such an evolutionary path is known as ``crossing a fitness valley".
Previous efforts at studying this problem have rigorously characterized the rate at which such complex adaptations evolve in populations of roughly equally fit individuals.
However, populations that are very large or have broad fitness distributions, such as many microbial populations, adapt quickly, which substantially alters the fate and dynamics of individual mutations due to the action of genetic draft.
We investigate the rate at which complex adaptations evolve in these rapidly adapting populations in regions without recombination.
We confirm that rapid adaptation overall increases the time required to cross a valley; however, rapid adaptation can make it easier for deeper valleys to be crossed relative to the time required for single beneficial mutations to sweep to fixation.
\end{abstract}

\maketitle

@@

The simplest adaptive scenarios in evolution involve the arisal and fixation of successive beneficial mutations.
This appears to be how Darwin thought even complex systems, such as the vertebrate eye, evolved \citep{darwin_1859}, and this assumption underlies many models in evolutionary theory including the those from adaptive dynamics \citep{Geritz:Kisdi:1998,Dercole:Rinaldi:2008} and theories of adaptation \citep{Gillespie:1983,Orr:1998,Gillespie:1991}.
Evolution will certainly proceed in this fashion if there always exists a sequence of mutations from an initial genotype to the high fitness genotype in which each mutant genotype in the sequence has a higher fitness than the previous one.
If the individual fitness of each genotype is visualized as a surface or landscape where the axes represent alternative alleles at each locus, then smooth landscapes with a single peak ensure that uphill paths exist no matter where a population starts.
However, many empirical fitness landscapes are not completely smooth and have multiple peaks \citep[reviewed in][]{Szendro:Schenk:2013,Visser:Krug:2014,Obolski:Ram:2018}.
In such landscapes, wild type individuals may have to traverse a mutational valley--a region of lower fitness--in order to reach a higher fitness peak.
This is a special case of ``sign epistasis".
We seek to characterize this ``valley crossing" process in order to better understand the routes adaptation is likely to take on these rugged empirical landscapes \citep[e.g.,][]{Aguilar-Rodriguez:Payne:2017} and to assess the likelihood of valley crossing under more complex scenarios in which it may in fact be common \citep[e.g.][]{trotter_2014}.

Valley crossing in asexual populations is increasingly well understood \citep{weissman_2009}.
If the number of mutants produced every generation is high, as is the case in large populations with high mutation rates, a complex adaptation requiring multiple mutations that is destined to fix will eventually be generated de novo.
This is sometimes called deterministic fixation of the multiple mutant\citep{weissman_2009}.
In small populations, neutral or deleterious intermediate mutants generated by wild type individuals can drift to fixation. Further mutations necessary for the adaptation can then arise on intermediate mutant backgrounds and fix due to positive selection.
This is the sequential fixation regime \citep{weissman_2009}.
A third possibility occurs for intermediate population sizes where wild type individuals generate transient mutant subpopulations or ``bubbles".
These bubbles are ordinarily doomed to extinction due to drift, but additional mutations in a lucky bubble may generate a complex adaptation that sweeps to fixation; this process has been referred to as ``tunneling" \citep{iwasa_2004, weissman_2009}.
On its own, valley crossing becomes more likely as the population size increases \citep{weissman_2009}. However, recent work suggests that valley crossing relative to the fixation of a simple beneficial mutation is least likely at intermediate population sizes where tunneling occurs \citet{ochs_2015}. This suggests that tunneling may be the most difficult mode of valley crossing. 
Recombination also affects the rate of valley crossing.
Infrequent recombination decreases valley crossing times \citep{weissman_2010}.
At high values of the recombination rate, however, individuals carrying the full complex adaptation outcross with the wild type and produce more deleterious intermediates that retard the growth of the multiple mutant.
An additional effect that can heighten the rate at which valley crossing occurs is population subdivision \citep{Bitbol:Schwab:2014}: the population size within a subdivision is smaller, meaning that selection against potentially harmful intermediates is relaxed.

@@

\begin{figure}
\begin{center}
  \includegraphics[scale=0.75]{Figures/valley.pdf}
  \caption{Fitness landscape.}
  \label{fig:landscape}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{subfigure}[b]{0.4\textwidth}
\includegraphics[width=\textwidth]{Figures/det_tunnel.pdf}
\end{subfigure}
\begin{subfigure}[b]{0.4\textwidth}
\includegraphics[width=\textwidth]{Figures/det_fix.pdf}
\end{subfigure}
\caption{Valley crossing times in the semi-deterministic tunneling regime as a function of $\sigma$. In both figures, $s = 10^{-2}$ and $\delta = 10^{-3}$: on the left, $N = 10^5$ and $\mu = 10^{-4}$, and on the right, $N = 10^6$ and $\mu = 10^{-3}$; the right figure has a high enough value of $N\mu^2$ that fixation is almost deterministic. Orange lines are averages, boxes are interquartile ranges, and end caps are overall ranges. Each box represents about $100$ simulation runs.}
\label{fig:deterministic}
\end{figure}

\begin{figure}
\begin{subfigure}[b]{0.4\textwidth}
\includegraphics[width=\textwidth]{Figures/seq_fix.pdf}
\end{subfigure}
\begin{subfigure}[b]{0.4\textwidth}
\includegraphics[width=\textwidth]{Figures/neut_tunnel.pdf}
\end{subfigure}
\caption{Valley crossing times in the sequential fixation (left) and tunneling (right) regimes as a function of $\sigma$. In both figures, $s = 10^{-2}$ and $\delta = 10^{-3}$: on the left, $N = 10^4$ and $\mu = 3\times 10^{-5}$, and on the right, $N = 10^5$ and $\mu = 3\times 10^{-6}$. Labeling is as in figure \ref{fig:deterministic}.}
\label{fig:tunneling}
\end{figure}

@@

\begin{figure}
\includegraphics[width=0.4\textwidth]{Figures/var_sigma_delta.pdf}
\caption{Valley crossing times for varied $\sigma$ and $\delta$. In these simulations, $N = 10^5$, $\mu = 3 \times 10^{-6}$, and $s = 10^{-2}$. Note the lack of dependence on $\delta$ at high $\sigma$.}
\label{fig:sigma_delta}
\end{figure}

@@

\begin{figure}
\begin{subfigure}[b]{0.475\textwidth}
  \includegraphics[width=\textwidth]{Figures/tau_compare_crop.pdf}
  \caption{Valley crossing times on a $\log_{10}$ scale}
  \label{fig:tau}
\end{subfigure}
\begin{subfigure}[b]{0.475\textwidth}
  \includegraphics[width=\textwidth]{Figures/alpha_ratios_crop.pdf}
  \caption{Ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$ on a $\log_{10}$ scale}
  \label{fig:ratios}
\end{subfigure}
\caption{Behavior of valley crossing times for varied $\delta$, $N$, and $\sigma$. In all cases, $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{valley}} = 0.1$, and $\mu = 10^{-5}$. Panel (a) shows that valley crossing times $\tau$ increase with $\delta$ and $\sigma$ but decrease with $N$, as predicted. Panel (b) shows the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$ as a function of $\delta$, $N$, and $\sigma$.}
\label{fig:tau-ratios}
\end{figure}

\begin{figure}
\includegraphics[width=0.95\textwidth]{Figures/compare_ratios_multi.pdf}
\caption{Ratio of $\alpha$ values at high and low $\sigma$ ($\log \left( \alpha_{\sigma_2} / \alpha_{\sigma_1} \right)$ where $\sigma_1=10^{-6}$ and $\sigma_2=10^{-2}$), for varied values of $\mu$. Genetic draft is helpful for crossing deep valleys in the sequential fixation and tunneling regimes, though at large population sizes it can slow down the crossing rate for deep valleys; at higher values of $\mu$ this limitation disappears.}
\label{fig:alpha_ratio}
\end{figure}

@@

@@

\clearpage
\bibliographystyle{genetics}
\bibliography{bib}

>>>>

\documentclass[10pt,twocolumn,twoside]{gsajnl}
\articletype{inv} % article type: {inv} Investigation

\usepackage{breqn}

\usepackage{subcaption}

@@

@@

\usepackage{setspace}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Genetic draft and valley crossing}
\author[$\ast$]{Taylor Kessinger}
\author[$\ast$,1]{Jeremy Van Cleve}
\affil[$\ast$]{Department of Biology, University of Kentucky}

\keywords{genetic drift; rapid adaptation; epistasis; fitness landscape}
\runningtitle{Genetic draft and valley crossing}
\runningauthor{Kessinger and Van Cleve}

\begin{abstract}
Living systems are characterized by complex adaptations which require multiple coordinated mutations in order to function. Empirical studies of fitness landscapes that result from the many possible mutations in a gene region reveal many fitness peaks and valleys that connect them. Thus, it is possible that some complex adaptations have arisen by evolutionary paths whose intermediate states are neutral or even deleterious.
When intermediates are deleterious, traversing such an evolutionary path is known as ``crossing a fitness valley".
Previous efforts at studying this problem have rigorously characterized the rate at which such complex adaptations evolve in populations of roughly equally fit individuals.
However, populations that are very large or have broad fitness distributions, such as many microbial populations, adapt quickly, which substantially alters the fate and dynamics of individual mutations due to the action of genetic draft.
We investigate the rate at which complex adaptations evolve in these rapidly adapting populations in regions without recombination.
We confirm that rapid adaptation overall increases the time required to cross a valley; however, rapid adaptation can make it easier for deeper valleys to be crossed relative to the time required for single beneficial mutations to sweep to fixation.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle
\thispagestyle{firststyle}
\marginmark
\firstpagefootnote

\correspondingauthoraffiliation{1}{Corresponding author: Department of Biology, University of Kentucky. E-mail: \href{emailto:jvancleve@uky.edu}{jvancleve@uky.edu}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The simplest adaptive scenarios in evolution involve the arisal and fixation of successive beneficial mutations.
This appears to be how Darwin thought even complex systems, such as the vertebrate eye, evolved \citep{darwin_1859}, and this assumption underlies many models in evolutionary theory including those from adaptive dynamics \citep{Geritz:Kisdi:1998,Dercole:Rinaldi:2008} and theories of adaptation \citep{Gillespie:1983,Orr:1998,Gillespie:1991}.
Evolution will certainly proceed in this fashion if there always exists a sequence of mutations from an initial genotype to the high fitness genotype in which each mutant genotype in the sequence has a higher fitness than the previous one.
If the individual fitness of each genotype is visualized as a surface or landscape where the axes represent alternative alleles at each locus, then smooth landscapes with a single peak ensure that uphill paths exist no matter where a population starts.
However, many empirical fitness landscapes are not completely smooth and have multiple peaks \citep[reviewed in][]{Szendro:Schenk:2013,Visser:Krug:2014,Obolski:Ram:2018}.
In such landscapes, wild type individuals may have to traverse a mutational valley--a region of lower fitness--in order to reach a higher fitness peak.
This is a special case of ``sign epistasis".
We seek to characterize this ``valley crossing" process in order to better understand the routes adaptation is likely to take on these rugged empirical landscapes \citep[e.g.,][]{Aguilar-Rodriguez:Payne:2017} and to assess the likelihood of valley crossing under more complex scenarios in which it may in fact be common \citep[e.g.][]{trotter_2014}.

Valley crossing in asexual populations is increasingly well understood \citep{weissman_2009}.
If the number of mutants produced every generation is high, as is the case in large populations with high mutation rates, a complex adaptation requiring multiple mutations that is destined to fix will eventually be generated de novo.
This is sometimes called deterministic fixation of the multiple mutant \citep{weissman_2009}.
In small populations, neutral or deleterious intermediate mutants generated by wild type individuals can drift to fixation. Further mutations necessary for the adaptation can then arise on intermediate mutant backgrounds and fix due to positive selection.
This is the sequential fixation regime \citep{weissman_2009}.
A third possibility occurs for intermediate population sizes where wild type individuals generate transient mutant subpopulations or ``bubbles".
These bubbles are ordinarily doomed to extinction due to drift, but additional mutations in a lucky bubble may generate a complex adaptation that sweeps to fixation; this process has been referred to as ``tunneling" \citep{iwasa_2004, weissman_2009}.
On its own, valley crossing becomes more likely as the population size increases \citep{weissman_2009}. However, recent work suggests that valley crossing relative to the fixation of a simple beneficial mutation is least likely at intermediate population sizes where tunneling occurs \citet{ochs_2015}. This suggests that tunneling may be the most difficult mode of valley crossing. 
Recombination also affects the rate of valley crossing.
Infrequent recombination decreases valley crossing times \citep{weissman_2010}.
At high values of the recombination rate, however, individuals carrying the full complex adaptation outcross with the wild type and produce more deleterious intermediates that retard the growth of the multiple mutant.
An additional effect that can heighten the rate at which valley crossing occurs is population subdivision \citep{Bitbol:Schwab:2014}: the population size within a subdivision is smaller, meaning that selection against potentially harmful intermediates is relaxed.

@@

\begin{figure}[t]
\begin{center}
  \includegraphics[scale=0.75]{Figures/valley.pdf}
  \caption{Fitness landscape.}
  \label{fig:landscape}
\end{center}
\end{figure}

@@

\begin{figure}[t]
\begin{subfigure}[b]{0.4\textwidth}
\includegraphics[width=\textwidth]{Figures/det_tunnel.pdf}
\end{subfigure}
\begin{subfigure}[b]{0.4\textwidth}
\includegraphics[width=\textwidth]{Figures/det_fix.pdf}
\end{subfigure}
\caption{Valley crossing times in the semi-deterministic tunneling regime as a function of $\sigma$. In both figures, $s = 10^{-2}$ and $\delta = 10^{-3}$: on the left, $N = 10^5$ and $\mu = 10^{-4}$, and on the right, $N = 10^6$ and $\mu = 10^{-3}$; the right figure has a high enough value of $N\mu^2$ that fixation is almost deterministic. Orange lines are averages, boxes are interquartile ranges, and end caps are overall ranges. Each box represents about $100$ simulation runs.}
\label{fig:deterministic}
\end{figure}

\begin{figure}[t]
\begin{subfigure}[b]{0.4\textwidth}
\includegraphics[width=\textwidth]{Figures/seq_fix.pdf}
\end{subfigure}
\begin{subfigure}[b]{0.4\textwidth}
\includegraphics[width=\textwidth]{Figures/neut_tunnel.pdf}
\end{subfigure}
\caption{Valley crossing times in the sequential fixation (left) and tunneling (right) regimes as a function of $\sigma$. In both figures, $s = 10^{-2}$ and $\delta = 10^{-3}$: on the left, $N = 10^4$ and $\mu = 3\times 10^{-5}$, and on the right, $N = 10^5$ and $\mu = 3\times 10^{-6}$. Labeling is as in figure \ref{fig:deterministic}.}
\label{fig:tunneling}
\end{figure}

@@

\begin{figure}[t]
\includegraphics[width=0.4\textwidth]{Figures/var_sigma_delta.pdf}
\caption{Valley crossing times for varied $\sigma$ and $\delta$. In these simulations, $N = 10^5$, $\mu = 3 \times 10^{-6}$, and $s = 10^{-2}$. Note the lack of dependence on $\delta$ at high $\sigma$.}
\label{fig:sigma_delta}
\end{figure}

@@

\begin{figure}[t]
\begin{subfigure}[b]{0.475\textwidth}
  \includegraphics[width=\textwidth]{Figures/tau_compare_crop.pdf}
  \caption{Valley crossing times on a $\log_{10}$ scale}
  \label{fig:tau}
\end{subfigure}
\begin{subfigure}[b]{0.475\textwidth}
  \includegraphics[width=\textwidth]{Figures/alpha_ratios_crop.pdf}
  \caption{Ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$ on a $\log_{10}$ scale}
  \label{fig:ratios}
\end{subfigure}
\caption{Behavior of valley crossing times for varied $\delta$, $N$, and $\sigma$. In all cases, $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{valley}} = 0.1$, and $\mu = 10^{-5}$. Panel (a) shows that valley crossing times $\tau$ increase with $\delta$ and $\sigma$ but decrease with $N$, as predicted. Panel (b) shows the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$ as a function of $\delta$, $N$, and $\sigma$.}
\label{fig:tau-ratios}
\end{figure}

@@

\begin{figure*}[t]
\includegraphics[width=0.95\textwidth]{Figures/compare_ratios_multi.pdf}
\caption{Ratio of $\alpha$ values at high and low $\sigma$ ($\log \left( \alpha_{\sigma_2} / \alpha_{\sigma_1} \right)$ where $\sigma_1=10^{-6}$ and $\sigma_2=10^{-2}$), for varied values of $\mu$. Genetic draft is helpful for crossing deep valleys in the sequential fixation and tunneling regimes, though at large population sizes it can slow down the crossing rate for deep valleys; at higher values of $\mu$ this limitation disappears.}
\label{fig:alpha_ratio}
\end{figure*}

@@

\bibliography{bib}
